An Evaluation of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-09-20 paper-submitted-3-nosubdir.tex
Updates from Overleaf

\title{A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates}

@@

Estimates of $a(\omega)$ and $b(\omega)$ in Model~2 are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS). At each evaluation frequency, a regression is performed using a set $G_E(\omega)$, $E_x(\omega)$, and $E_y(\omega)$ values from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data in a band of frequencies around $\omega_e$. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}.

@@

Out-of-sample estimation quality metrics for each model on day $D$ were determined by computing the average model coefficients using the other 34 days of data and then using these average coefficients to generate GIC time series and associated metrics on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model. In calculating the metrics, the first and last 10 minutes of the day were excluded to reduce the influence of transients that exist in the estimate at the start of the interval for the amount of time that the causal part of the transfer function impulse responses (computed by Fourier inversion of the interpolated transfer function components) is non-zero and at the end for the amount of time that the acausal part of the impulse responses are non-zero.

@@

(3) The coefficients in Model~1 are determined assuming that the electric field is spatially uniform at all points in the system. The exact calculation of GICs in a power system under the quasi-DC assumption requires knowledge of the electric field at all points of the system, and simulations of GIC flow typically start with an assumed electric field along the lines to calculate input voltages \citep{Albertson1981,Lehtinen1985, Pirjola2009}. The electric fields are then used along with the DC system component resistances and their connectivity to calculate GIC in the different parts of the system. In both Model~1 and Model~2, the input electric field is typically assumed to be spatially uniform along the line in which GIC was measured so that the potential difference is directly proportional to the electric field. Spatial variations in the electric field measured on Earth's surface can be due to horizontal variations in the incident electric field and horizontal variations in sub-surface conductivity, including those at land-ocean interfaces. If a spatial variation exists along the line in which GIC was measured, the coefficients derived from measurements in both models may differ significantly from those calculated using system configuration and resistance information. However, the data-derived coefficients will not have a frequency dependence that results only from the variation in the direction of a static electric field along the line.

@@

The primary implications of the results presented are (1) significantly improved estimates of GICs using data-derived system coefficients $a$ and $b$ can be made if they are allowed to be frequency dependent and (2) a direct comparison of data-derived $a$ and $b$ coefficients and those computed using power system configuration information will be problematic if the data are not consistent with the assumptions of Equation~\ref{model1}.

@@

\acknowledgments
Magnetic and electric field measurements made at the Kakioka Magnetic Observatory were obtained from the Japan Meteorological Agency Data Portal \url{http://www.kakioka-jma.go.jp/obsdata/metadata/en}. The geomagnetic coordinates in Figure~\ref{map} were computed using code from \url{https://www.ngdc.noaa.gov/geomag/geom_util/apex.shtml}. The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]. We thank the Hokkaido Electric Power Co. for the GIC measurements at the Memanbetsu substation and Shinichi Watari for communicating and preparing the GIC data and providing feedback on the manuscript. We also thank Ikuko Fujii for providing clarifications on the geoelectric instrument response, Trevor Gaunt and Emanuel Benabeau for review and comments on part (1) of Section~\ref{discussion}, and Josh Rigler for substantial feedback and review.
%\clearpage

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/map.pdf}
  \caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memanbetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the Memanbetsu substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
  \label{map}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_raw_All_20060805-v1.pdf}
  \caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB). (Bottom) GIC measurements at Memanbetsu substation. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, the date labels are for Universal Time, and the average has been subtracted from each time series.}
  \label{sample}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_GIC_predictions-MeanModel-2006-08-06-v0-o0.pdf}
  \caption{Out-of-sample model predictions (top) and predictions errors for each model in a selected 1-day interval. The scales for the error time series differ from that for the top panel so that the error amplitudes are scaled up by a factor of 2.}
  \label{predictions}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_SN-options-1-v0-o0.pdf}
  \caption{Out-of-sample signal-to-noise (signal to prediction error) ratios at the evaluation frequencies for the four models.}
  \label{SN}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_aobo_histograms-options-1-v0-o0.pdf}
  \caption{Distribution of coefficients in Model~1. Each set of coefficients was computed 35 times using one day of 1-second cadence data. The circle or dot on the horizontal axis indicates the average of a histogram and $\overline{a}_o$ = 73 A/(V/km) and $\overline{b}_o$ = -72 A/(V/km).}
  \label{histogram}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Z-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the coefficients in Models 1, 2, and 4. The frequency domain transfer functions for $a_o$ and $b_o$ in Model~1 are constant and equal to $a_o$ and $b_o$.}
  \label{Z}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Phi-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the coefficients in Models 1, 2, and 4. The phase for $a_o$ and $b_o$ are constant and either 0$^{\circ}$ or 180$^{\circ}$ depending on the sign of their value, with positive values having a phase of 0$^{\circ}$.}
  \label{Phi}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Z_MT-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Z_MT}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Phi_MT-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Phi_MT}
\end{figure}

>>>>

\title{An Evaluation of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates}

@@

Estimates of $a(\omega)$ and $b(\omega)$ in Model~2 are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing an ordinary least squares regression (OLS). At each evaluation frequency, a regression is performed using a set $G_E(\omega)$, $E_x(\omega)$, and $E_y(\omega)$ values from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data in a band of frequencies around $\omega_e$. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}.

@@

Out-of-sample estimation quality metrics for each model on day $D$ were determined by computing the average model coefficients using the other 34 days of data and then using these average coefficients to generate GIC time series and associated metrics on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model. In calculating the metrics, the first and last 10 minutes of the day were excluded to reduce the influence of transients at the start and end of the interval.

@@

(3) The coefficients in Model~1 are determined assuming that the electric field is spatially uniform at all points in the system. The exact calculation of GICs in a power system under the quasi-DC assumption requires knowledge of the electric field at all points of the system, and simulations of GIC flow typically start with an assumed electric field along the lines to calculate input voltages \citep{Albertson1981,Lehtinen1985, Pirjola2009}. The electric fields are then used along with the DC system component resistances and their connectivity to calculate GIC in the different parts of the system. In both Model~1 and Model~2, the input electric field is typically assumed to be spatially uniform along the line in which GIC was measured so that the induced voltage is directly proportional to the electric field. Spatial variations in the electric field measured on Earth's surface can be due to horizontal variations in the incident electric field and horizontal variations in sub-surface conductivity, including those at land-ocean interfaces. If a spatial variation exists along the line in which GIC was measured, the coefficients derived from measurements in both models may differ significantly from those calculated using system configuration and resistance information. However, the data-derived coefficients will not have a frequency dependence that results only from the variation in the direction of a static electric field along the line.

@@

The primary implications of the results presented are (1) significantly improved estimates of GICs using data-derived system coefficients $a$ and $b$ can be made if they are allowed to be frequency dependent and (2) $a$ and $b$ coefficients derived using a traditional frequency-independent model may differ significantly from those computed using power system configuration information.

%a direct comparison of data-derived $a$ and $b$ coefficients and those computed using power system configuration information will be problematic if the data are not consistent with the frequency independent assumption.

@@

\acknowledgments
Magnetic and electric field measurements made at the Kakioka Magnetic Observatory were obtained from the Japan Meteorological Agency Data Portal \url{http://www.kakioka-jma.go.jp/obsdata/metadata/en}. The geomagnetic coordinates in Figure~\ref{map} were computed using code from \url{https://www.ngdc.noaa.gov/geomag/geom_util/apex.shtml}. The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]. We thank the Hokkaido Electric Power Co. for the GIC measurements at the Memanbetsu substation and Shinichi Watari for communicating and preparing the GIC data and providing feedback on the manuscript. We also thank Ikuko Fujii for providing clarifications on the geoelectric instrument response, Trevor Gaunt and Emanuel Bernabeau for review and comments on part (1) of Section~\ref{discussion}, and Josh Rigler for substantial feedback and review.
%\clearpage

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{map.pdf}
  \caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memanbetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the Memanbetsu substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
  \label{map}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_raw_All_20060805-v1.pdf}
  \caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB). (Bottom) GIC measurements at Memanbetsu substation. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, the date labels are for Universal Time, and the average has been subtracted from each time series.}
  \label{sample}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_GIC_predictions-MeanModel-2006-08-06-v0-o0.pdf}
  \caption{Out-of-sample model predictions (top) and predictions errors for each model in a selected 1-day interval. The scales for the error time series differ from that for the top panel so that the error amplitudes are scaled up by a factor of 2.}
  \label{predictions}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_SN-options-1-v0-o0.pdf}
  \caption{Out-of-sample signal-to-noise (signal to prediction error) ratios at the evaluation frequencies for the four models.}
  \label{SN}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_aobo_histograms-options-1-v0-o0.pdf}
  \caption{Distribution of coefficients in Model~1. Each set of coefficients was computed 35 times using one day of 1-second cadence data. The circle or dot on the horizontal axis indicates the average of a histogram and $\overline{a}_o$ = 73 A/(V/km) and $\overline{b}_o$ = -72 A/(V/km).}
  \label{histogram}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Z-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the coefficients in Models 1, 2, and 4. The frequency domain transfer functions for $a_o$ and $b_o$ in Model~1 are constant and equal to $a_o$ and $b_o$.}
  \label{Z}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Phi-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the coefficients in Models 1, 2, and 4. The phase for $a_o$ and $b_o$ are constant and either 0$^{\circ}$ or 180$^{\circ}$ depending on the sign of their value, with positive values having a phase of 0$^{\circ}$.}
  \label{Phi}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Z_MT-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Z_MT}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Phi_MT-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Phi_MT}
\end{figure}

### 2019-09-20 paper-submitted-3.tex
Updates from Overleaf

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{map.pdf}
  \caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memanbetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the Memanbetsu substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
  \label{map}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_raw_All_20060805-v1.pdf}
  \caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB). (Bottom) GIC measurements at Memanbetsu substation. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, the date labels are for Universal Time, and the average has been subtracted from each time series.}
  \label{sample}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_GIC_predictions-MeanModel-2006-08-06-v0-o0.pdf}
  \caption{Out-of-sample model predictions (top) and predictions errors for each model in a selected 1-day interval. The scales for the error time series differ from that for the top panel so that the error amplitudes are scaled up by a factor of 2.}
  \label{predictions}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_SN-options-1-v0-o0.pdf}
  \caption{Out-of-sample signal-to-noise (signal to prediction error) ratios at the evaluation frequencies for the four models.}
  \label{SN}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_aobo_histograms-options-1-v0-o0.pdf}
  \caption{Distribution of coefficients in Model~1. Each set of coefficients was computed 35 times using one day of 1-second cadence data. The circle or dot on the horizontal axis indicates the average of a histogram and $\overline{a}_o$ = 73 A/(V/km) and $\overline{b}_o$ = -72 A/(V/km).}
  \label{histogram}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Z-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the coefficients in Models 1, 2, and 4. The frequency domain transfer functions for $a_o$ and $b_o$ in Model~1 are constant and equal to $a_o$ and $b_o$.}
  \label{Z}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Phi-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the coefficients in Models 1, 2, and 4. The phase for $a_o$ and $b_o$ are constant and either 0$^{\circ}$ or 180$^{\circ}$ depending on the sign of their value, with positive values having a phase of 0$^{\circ}$.}
  \label{Phi}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Z_MT-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Z_MT}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{plot_model_summary_Phi_MT-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Phi_MT}
\end{figure}

>>>>

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/map.pdf}
  \caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memanbetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the Memanbetsu substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
  \label{map}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_raw_All_20060805-v1.pdf}
  \caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB). (Bottom) GIC measurements at Memanbetsu substation. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, the date labels are for Universal Time, and the average has been subtracted from each time series.}
  \label{sample}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_GIC_predictions-MeanModel-2006-08-06-v0-o0.pdf}
  \caption{Out-of-sample model predictions (top) and predictions errors for each model in a selected 1-day interval. The scales for the error time series differ from that for the top panel so that the error amplitudes are scaled up by a factor of 2.}
  \label{predictions}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_SN-options-1-v0-o0.pdf}
  \caption{Out-of-sample signal-to-noise (signal to prediction error) ratios at the evaluation frequencies for the four models.}
  \label{SN}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_aobo_histograms-options-1-v0-o0.pdf}
  \caption{Distribution of coefficients in Model~1. Each set of coefficients was computed 35 times using one day of 1-second cadence data. The circle or dot on the horizontal axis indicates the average of a histogram and $\overline{a}_o$ = 73 A/(V/km) and $\overline{b}_o$ = -72 A/(V/km).}
  \label{histogram}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Z-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the coefficients in Models 1, 2, and 4. The frequency domain transfer functions for $a_o$ and $b_o$ in Model~1 are constant and equal to $a_o$ and $b_o$.}
  \label{Z}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Phi-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the coefficients in Models 1, 2, and 4. The phase for $a_o$ and $b_o$ are constant and either 0$^{\circ}$ or 180$^{\circ}$ depending on the sign of their value, with positive values having a phase of 0$^{\circ}$.}
  \label{Phi}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Z_MT-options-1-v0-o0.pdf}
  \caption{Frequency domain transfer functions for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Z_MT}
  %\end{figure}

@@

  %\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_model_summary_Phi_MT-options-1-v1-o0.pdf}
  \caption{Frequency domain phase values for the $\mathbf{Z}$ coefficients used in Model~3.}
  \label{Phi_MT}
\end{figure}
